### 2024-02-12 parts/correction.tex
big big rewrite

\begin{corollary}
    \label{n-rat-correct:lem}
    Let $P \in \Rel[X_1, \dots, X_p]$ be \kl{represented}
    by a \kl{$\Nat$-rational series}. Then,
    $P \in \CorrectPoly$.
\end{corollary}
\begin{proof}
    Let $f$ be the \kl{commutative}
    \kl{$\Nat$-rational series}
    with domain $\Sigma \defined \set{a_1, \dots, a_p}$
    that \kl{represents} $P$. Because $f$ has \kl{polynomial growth},
    it is in fact a \kl{$\Nat$-polyregular function}
    (\cref{polynomial-rational-polyreg:fact}).
    Using \cref{n-poly-combinatorics:lem},
    there exists a number $\omega \in \Nat$
    and polynomial $Q$ with \kl{non-negative}
    \kl{maximal monomials} such that
    for all $X_1, \dots, X_p \geq \omega$:
    \begin{equation*}
        f\left(
            \prod_{i = 1}^p a_i^{\omega X_i}
        \right)
        = Q(X_1, \dots, X_p)
        = P(\omega X_1, \dots, \omega X_p) 
        \quad .
    \end{equation*}
    We conclude that $P(\omega X_1, \dots, \omega X_p) = Q(X_1, \dots, X_p)$
    as polynomials,
    and in particular that
    the \kl{maximal monomials} of 
    $P$ are \kl{non-negative}.

    Remark that for every partial valuation $\nu \colon \vec{X} \topartial \Nat$,
    the polynomial $\restr{P}{\nu}$ continues to be represented
    by a \kl{$\Nat$-polyregular function}, namely
    $f$ partially applied to a word. As a consequence,
    the \kl{maximal monomials} of
    $\restr{P}{\nu}$ are also \kl{non-negative}, 
    and
    we have proven that $P \in \CorrectPoly$.
\end{proof}

@@

As one would expect, the \kl{discrete derivatives} are linear operations on
polynomials, that commutes with \kl{translation operators}. However, the
\kl{translations operators} do not perfectly commute with partial applications
$\restr{\cdot}{\nu}$.

\begin{fact}
    \label{discrete-deriv-linear:fact}
    For all tuples $\vec{X}$ of indeterminates,
    for all $K \in \Nat$, $L \in \Rel$, for all partial function
    $\nu \colon \vec{X} \topartial \Nat$:
    \begin{enumerate}
        \item The maps $\translate{K} \colon \Rel[\vec{X}] \to \Rel[\vec{X}]$,
        $\Diff{K}{ \cdot } \colon \Rel[\vec{X}] \to \Rel[\vec{X}]$,
        and
        $\restr{\cdot}{\nu} \colon \Rel[\vec{X}] \to \Rel[\vec{X}]$
            are linear operators,
        \item $\Diff{K}{ \cdot } \circ \translate{L}
            = \translate{K} \circ \Diff{L}{\cdot}$,
        \item $\restr{\cdot}{\nu} \circ \translate{L}
            = \translate{L} \circ \restr{\cdot}{\translate{L}(\nu)}$.
    \end{enumerate}
\end{fact}

To build some intuition on the proof to come, let us sketch the argument in the
case of a single indeterminate $X$. Given a polynomial $P \in \CorrectPoly$,
the goal is to build a \kl{$\Nat$-polyregular function} $f$ that
\kl{represents} $P$. Note that in this case, the function has unary input and
unary output. We prove the result by induction on the degree $\deg(P)$ of the
polynomial $P$, where the base case is trivial because when $P$ is a constant
function, this constant must be in $\Nat$. For the induction step, we can
assume that $P = \alpha X^n + Q$, where $\deg(Q) < \deg(P)$, $n \geq 1$, and
$\alpha > 0$. On the one hand, $\alpha X^n$ is a \kl{$\Nat$-rational
polynomial}, but on the other hand, $Q$ may not belong to $\CorrectPoly$.
However, there exists $K \in \Nat$ such that $\Diff{K}{X^n} + \translate{K}(Q)$
belongs to $\CorrectPoly$, for instance by considering $K$ to be the sum of
absolute values of coefficients of $P$. Furthermore, $\Diff{K}{X^n} +
\translate{K}(Q)$ has degree smaller than $\deg(P)$, hence is a
\kl{$\Nat$-rational polynomial}. Now, $\translate{K}(P) = \alpha X^n +
\Diff{K}{X^n} + \translate{K}(Q)$ is a \kl{$\Nat$-rational polynomial},
represented by a function $f_1$. Now, for values $k$ of $X$ that are smaller
than $K$, we can compute $P(k) = \restr{P}{X=k}$, which belongs to $\Nat$. It
is possible to build a \kl{$\Nat$-polyregular function} $f_2 \colon \Nat \to
\Nat$ that outputs $P(k)$ if the input is at most $K$, and $f_1(k - K)$
otherwise.

@@

When multiple indeterminate are involved, the notion of 
degree must be adapted, and to ensure that the decision procedure
is effective, one must obtain explicit bounds on the
translations that are needed.

\AP To that end, let us now introduce the ordering over which the induction of
\cref{lem:correct-to-n-rat} is built. Recall that $\MaximalMonomials(P)$ is the
set of \kl{maximal monomials} of $P$, hence belongs to $\Pfin(\Monomials)$. We
order \kl{monomials} with the \kl{divisibility ordering}, making $\Monomials$ a
\kl{well-quasi-ordering} that is isomorphic to $\Nat^k$ with the product
ordering \cite[see e.g.][Dickson’s Lemma]{SCSC12}. We endow $\Pfin(\Monomials)$
with the \intro{Hoare ordering}, that is, $S_1 \hoareleq S_2$ whenever for all
\kl{monomials} $M_1 \in S_1$, there exists a monomial $M_2 \in S_2$, such that
$M_1$ \kl{divides} $M_2$. The set $(\Pfin(\Monomials), \hoareleq)$ remains a
\kl{well-quasi-ordering} \cite[see e.g.][Hoare quasi-ordering]{SCSC12}.

\begin{fact}
    \label{maxi-monomials-submodular:fact}
    Let $P, Q$ be two polynomials.
    Then, $\MaximalMonomials(P+Q) \hoareleq \MaximalMonomials(P)
    \cup \MaximalMonomials(Q)$.
\end{fact}

@@

\begin{fact}
    \label{translation-maximal:fact}
    For all $P \in \Rel[\vec{X}]$, and $K \in \Nat$,
    \begin{equation*}
        \MaximalMonomials(P) = \MaximalMonomials(\translate{K}(P))
        \quad .
    \end{equation*}
\end{fact}

@@

    For all partial valuations $\nu \colon \vec{X} \topartial \set{0, \dots,
    K}$ fixing at least one indeterminate, $\MaximalMonomials(\restr{P}{\nu})
    \hoarele \MaximalMonomials{P}$. Because $\restr{P}{\nu} \in \CorrectPoly$,
    one can use the induction hypothesis to compute a \kl{star-free
    $\Nat$-polyregular function} $f_\nu$ that \kl{represents} $\restr{P}{\nu}$.

@@

\begin{corollary}
    \label{decidability-correct:lem}
    One can decide membership in $\CorrectPoly$.
\end{corollary}
\begin{proof}
    Given $P \in \Rel[\vec{X}]$,
    one applies \cref{derivation-translation:lem}
    to obtain a bound $K$, and then 
    computes $\translate{K}(\restr{P}{\nu})$ for all 
    partial valuations $\nu \colon \vec{X} \topartial \set{0, \dots, K}$.
    Then $P \in \CorrectPoly$ if and only if all those polynomials
    belong to $\Nat[\vec{X}]$, which is decidable.
\end{proof}

>>>>

\begin{corollary}
    \label{n-rat-correct:lem}
    Let $P \in \Rel[X_1, \dots, X_p]$ be \kl{represented}
    by a \kl{$\Nat$-rational series}. Then,
    $P \in \CorrectPoly$.
\end{corollary}

@@

As one would expect, the \kl{discrete derivatives} are linear operations on
polynomials, that commutes with \kl{translation operators}. However, the
\kl{translation operators} do not perfectly commute with partial applications
$\restr{\cdot}{\nu}$.

\begin{fact}
    \label{discrete-deriv-linear:fact}
    For all tuples $\vec{X}$ of indeterminates,
    for all $K \in \Nat$, $L \in \Rel$, for all partial function
    $\nu \colon \vec{X} \topartial \Nat$:
    \begin{enumerate}
        \item The maps $\translate{K} \colon \Rel[\vec{X}] \to \Rel[\vec{X}]$,
        $\Diff{K}{ \cdot } \colon \Rel[\vec{X}] \to \Rel[\vec{X}]$,
        and
        $\restr{\cdot}{\nu} \colon \Rel[\vec{X}] \to \Rel[\vec{X}]$
            are linear operators,
        \item $\Diff{K}{ \cdot } \circ \translate{L}
            = \translate{L} \circ \Diff{K}{\cdot}$,
        \item $\restr{\cdot}{\nu} \circ \translate{L}
            = \translate{L} \circ \restr{\cdot}{\translate{L}(\nu)}$.
    \end{enumerate}
\end{fact}

@@

\AP When multiple indeterminate are involved, the notion of degree must be
adapted, and to ensure that the decision procedure is effective, one must
obtain explicit bounds on the translations that are needed. To that end, let us
now introduce the ordering over which the induction of
\cref{lem:correct-to-n-rat} is built. Recall that $\MaximalMonomials(P)$ is the
set of \kl{maximal monomials} of $P$, hence belongs to $\Pfin(\Monomials)$. We
order \kl{monomials} with the \kl{divisibility ordering}, making $\Monomials$ a
\kl{well-quasi-ordering} that is isomorphic to $\Nat^k$ with the product
ordering \cite[Dickson’s Lemma]{SCSC12}. We endow $\Pfin(\Monomials)$ with the
\intro{Hoare ordering}, that is, $S_1 \hoareleq S_2$ whenever for all
\kl{monomials} $M_1 \in S_1$, there exists a monomial $M_2 \in S_2$, such that
$M_1$ \kl{divides} $M_2$. The set $(\Pfin(\Monomials), \hoareleq)$ remains a
\kl{well-quasi-ordering} \cite[Hoare quasi-ordering]{SCSC12}.

@@

@@

    For all partial valuations $\nu \colon \vec{X} \topartial \set{0, \dots,
    K}$ fixing at least one indeterminate,  we know that
    $\MaximalMonomials(\restr{P}{\nu}) \hoarele \MaximalMonomials{P}$. Because
    $\restr{P}{\nu} \in \CorrectPoly$, one can use the induction hypothesis to
    compute a \kl{star-free $\Nat$-polyregular function} $f_\nu$ that
    \kl{represents} $\restr{P}{\nu}$.

@@

\begin{corollary}
    \label{decidability-correct:lem}
    One can decide membership in $\CorrectPoly$.
\end{corollary}

### 2024-02-12 parts/counter.tex
big big rewrite

%! TeX program = xelatex
%! lang = en-US
\section{Polynomials}
\label{polynomials:sec}

The goal of this section is to provide semantic characterizations of
\kl{$\Nat$-rational polynomials}, decide if a polynomial $P$ is a
\kl{$\Nat$-rational polynomial}, and if so, explicitly compute a
\kl{$\Nat$-polyregular function} $f$ that \kl{represents} $P$.

It is an easy check that polynomials with coefficients in $\Nat$ are
\kl{$\Nat$-rational polynomials} (\cref{n-poly-n-poly:example}). However,
\cref{negative-but-npoly:ex} provides an example of a polynomial with negative
coefficients that is a \kl{$\Nat$-rational polynomial}. As a consequence, the
class of interest strictly contains $\Nat[\vec{X}]$.

@@

\begin{example}
    \label{negative-but-npoly:ex}
    Let $P \defined X^2 -2X + 2$. Then $P$
    is a \kl{$\Nat$-rational polynomial}.
\end{example}
\begin{proof}
    Remark that $P = (X-1)^2 + 1$. Let $f  \colon \Sigma^* \to \Nat$
    that maps $aw \mapsto |w|^2$, and $\varepsilon \to 0$.
    It is clear that $f \in \NPoly$, and that 
    $P$ is \kl{represented} by
    $\ind{|w| \geq 1} \times f + \ind{|w| = 0}$.
\end{proof}

@@

Let us now introduce the counter example that was found trying to generalize
the original proof $\ZPoly$.
The counter example will use three indeterminates, and
we will prove in \cref{sec:proof} that \citeauthor{KARH77}’s theorem holds on
polynomial with at most two indeterminate. In particular, the examples
appearing in \cite{KARH77} are not invalidated, as they all use at most two
indeterminates. Furthermore, the converse implication, that is, that
\kl{$\Nat$-rational polynomials} belong to $\CoveredPoly$ is correct, and can
even be strengthened as we will see in \cref{corrected-version:thm}.

@@

\begin{lemma}
    \label{thm:counter-example}
    The polynomial $\BadPoly$ belongs to $\CoveredPoly$,
    but is not a \kl{$\Nat$-rational polynomial}.
\end{lemma}
\begin{proof}
    It is clear that $\BadPoly$ is \kl{non-negative}. We can develop
    the expression of $\BadPoly$ to 
    obtain
    $\BadPoly = ZX^2 + ZY^2 + 2ZXY + 2X^2 -4XY + 2Y^2$.
    The \kl{maximal monomials} of $P$
    are $ZX^2$, $ZY^2$, and $2ZXY$, all of which have
    \kl{positive coefficients}.

>>>>

%! TeX program = xelatex
%! lang = en-US
\section{$\Nat$-rational Polynomials}
\label{polynomials:sec}

The goal of this section is to provide semantic characterizations of
\kl{$\Nat$-rational polynomials}, decide if a polynomial $P$ is a
\kl{$\Nat$-rational polynomial}, and if so, explicitly compute a
\kl{$\Nat$-polyregular function} $f$ that \kl{represents} $P$. It is an easy
check that polynomials with coefficients in $\Nat$ are \kl{$\Nat$-rational
polynomials} (\cref{n-poly-n-poly:example}). However,
\cref{negative-but-npoly:ex} provides a polynomial with negative coefficients
that is a \kl{$\Nat$-rational polynomial}. As a consequence, the class of
interest strictly contains $\Nat[\vec{X}]$.

@@

\begin{example}
    \label{negative-but-npoly:ex}
    Let $P \defined X^2 -2X + 2$. Then $P$
    is a \kl{$\Nat$-rational polynomial}.
\end{example}

@@

Let us now provide a counter example to \cref{karh:thm}. The counter example
will use three indeterminates, and we will prove in \cref{sec:proof} that
\cref{karh:thm} holds on polynomial with at most two indeterminate. In
particular, the examples appearing in \cite{KARH77} are not invalidated, as
they all use at most two indeterminates. Furthermore, the converse implication,
that is, that \kl{$\Nat$-rational polynomials} belong to $\CoveredPoly$ is
correct, and can even be strengthened as we will see in
\cref{corrected-version:thm}.

@@

\begin{lemma}
    \label{thm:counter-example}
    The polynomial $\BadPoly$ belongs to $\CoveredPoly$,
    but is not a \kl{$\Nat$-rational polynomial}.
\end{lemma}
\begin{proof}
    It is clear that $\BadPoly$ is \kl{non-negative}. We can develop
    the expression of $\BadPoly$ to 
    obtain
    $\BadPoly = ZX^2 + ZY^2 + 2ZXY + 2X^2 -4XY + 2Y^2$.
    The \kl{maximal monomials} of $P$
    are $ZX^2$, $ZY^2$, and $2ZXY$, all of which are
    \kl{non-negative}.

### 2024-02-12 parts/intro.tex
adding polyblind?
test
big big rewrite

There have been recurrent attempts to generalize the notions of regularity from
languages to functions, leading to classes of various expressiveness
(sequential, rational, regular, and \kl{polyregular functions}). In this paper,
we will focus on \kl{polyregular functions} ($\Poly$), that shares with regular
languages its characterizations in terms of logic (via $\MSO$ interpretations),
automata (via pebble automata), and adds new bridges with programming languages
via its relationships with lambda-terms and so-called for-transducers
\cite[see, e.g.][]{bojanczyk2019string}. As for regular languages, the class
$\SF$ of \intro{star-free polyregular function} can equivalently be introduced
in terms of first-order definability, counter-free pebble transducers, and
suitable restrictions of for-tranducers.

Classical problems, including the decidability of aperiodicity, become
difficult in the case of \kl{polyregular functions}, due to the lack of
canonical objects akin to the syntactic monoid or the minimal automaton.
Indeed, for sequential functions that admit a canonical object, aperiodicity is
decidable \cite{choffrut03}, and more recently, aperiodicity has been proven
decidable for rational functions
\cite{filiot2016aperiodicity,filiot2018canonical}. The problem remains open for
regular functions, although it has been proven decidable when changing the
semantics of function equality to account for the \emph{origins} in the output
\cite{bojanczyk14}. To construct subclasses of \kl{polyregular function}, an
interesting parameter is the \kl{growth rate} of a function $f$, which is given
by a number $k \in \Nat$, which can be used to create a strict hierarchy inside
$\Poly$. We write $\Poly[k]$ for the \kl{polyregular functions} of \kl{growth
rate} $k$. It has been shown that the parameter $k$ controls the maximal
complexity of the $\MSO$ interpretation needed to represent the function
\cite{bojanczyk2022transducers}.

@@

Recently, aperiodicity has been shown decidable over a restricted class of
\kl{polyregular functions} where the ouptut belongs to $\Rel$ instead of
$\Nat$, that are called \kl{$\Rel$-polyregular functions} ($\ZPoly$)
\cite{LOPEZ23b}. These functions generalize \kl{$\Nat$-polyregular functions}
by allowing negative outputs, and therefore enabling a limited form of
backtracking. As for the unary outputs, these results generalize to
\kl{polyregular functions} having outputs in the free commutative \emph{group}
generated by a finite alphabet $\Gamma$. The decidability of aperiodicity for
\kl{$\Rel$-polyregular functions} relies on the construction of a canonical
object called the \emph{residual transducer}, the latter being essentially
based on differences between functions, crucially leveraging negative outputs
(i.e., the presence of inverse elements). If the proof method does not carry
from $\Rel$-output to $\Nat$-output functions, it was conjectured that the
semantic characterization of \kl{star-free $\Rel$-polyregular functions}
($\ZSF$) would continue to characterize \kl{star-free $\Nat$-polyregular
functions} ($\NSF$).

@@

It is well known that $\NPoly$ (resp. $\ZPoly$) is proper subclass respectively
of \kl{$\Nat$-rational series} ($\NRat$) (resp. of \kl{$\Rel$-rational series}
$\ZRat$), that have been characterized in terms of \kl{growth rate}, ambiguity,
and a notion of \emph{quantitative} $\MSO$
\cite{schutzenberger1962,kreutzer2013}, and therefore can benefit from generic
theorems obtained on these \cite{berstel2011noncommutative}. However, for these
larger classes of functions, several competing notions of aperiodicity exist
\cite{droste2019aperiodic,reutenauer_series_1980,LOPEZ23b}, and the
decidability of the ones that extends $\ZSF$ and $\NSF$ remains open.
Furthermore, the decidability of $\NRat$ inside $\ZRat$ is also a longstanding
open problem, that was to the knowledge of the authors, only solved by
\textcite{KARH77} in the case of \kl{$\Rel$-rational series} that compute
\emph{polynomials}. The decidability of $\NPoly$ inside $\ZPoly$ (resp. $\NSF$
inside $\ZSF$) remains an open problem
\cite[see e.g.][Open question 5.55]{gaetanphd}.

@@

\begin{figure}
    \includestandalone[width=\linewidth]{tikz/class-inclusions}
    \caption{
        Decidability and inclusions of classes of functions,
        arranged along two axes. The first one is the complexity
        of the output alphabet ($\Nat$, $\Rel$, $\Sigma$). The second
        one is the allowed computational power
        (star-free polyregular functions, polyregular functions, 
        rational series).
    }
    \label{previously-known-inclusions:fig}
\end{figure}

\paragraph{Contributions.} This paper focuses on \kl{$\Nat$-polyregular
functions} that are furthermore \kl{commutative}. Remark that this enforces a
commutativity property both on the input alphabet and on the output alphabet.
In this restricted setting, we answer positively to \cref{zsf-nsf:conjecture},
and prove that the membership problems between $\NSF$, $\ZSF$, $\NPoly$, and
$\ZPoly$ are all decidable, with effective conversion algorithms. This
effectively shows that all arrows in the lower left square of
\cref{previously-known-inclusions:fig} are decidable in the \kl{commutative}
setting. Furthermore, we characterize \kl{commutative} \kl{$\Nat$-polyregular
function} as the class of \kl{$\Rel$-polyregular functions} that are
\kl{combinatorial} (defined page \kpageref{combinatorial}), which is a semantic
property akin to the one \kl{ultimate polynomiality} that characterizes $\ZSF$.
As a byproduct of our analysis, we provide a counter example to the
characterization of polynomials computed by \kl{$\Nat$-rational series} given
by \textcite[Theorem 3.3, page 4; re-introduced as \cref{karh:thm}]{KARH77}.

Finally, we make a first step towards understanding the non-commutative by
introducing a canonical model of computation for \kl{$\Nat$-polyregular
functions} built upon residuals, called the \kl{residual transducer}. This
allows the characterization of $\NPoly[k]$ as the class of functions $f$ such
that the collection $\Res(f)$ of all \kl{residuals} of $f$ is a
\kl{well-quasi-ordering} with respect to the partial ordering $g \npolyleq[k]
h$ if and only if $h - g \in \NPoly[k-1]$. In this sense, it generalizes to
orderings the characterization of $\ZPoly[k]$ as the class of functions $f$
such that $\Res(f)$ has finite index for the equivalence relation $g
\zpolyequiv[k] h$ if and only if $g - h \in \ZPoly[k-1]$. The \kl{residual
transducer} is effectively computable if $f$ is a \kl{commutative}
\kl{$\Nat$-polyregular function}, and it is conjectured to be computable for
all functions in $\NPoly$. Finally, we prove that for $f \in \NPoly[0]$, $f \in
\NSF$ if and only if the \kl{residual transducer} of $f$ has no \kl{counters},
and conjecture that in full generality, $f \in \NSF$ if and only if the
\kl{residual transducer} of $f$ is \kl{counter-free}.

\paragraph{Outline of the paper.} In \cref{preliminaries:sec}, we introduce the
multiple characterizations of \kl{$\Nat$-polyregular function} (resp.
$\Rel$-polyregular functions), show that one can decide if a function $f \in
\ZPoly$ is \kl{commutative} (\cref{decide-commutative:lemma}), and give the
basic definitions used on multivariate polynomials. In \cref{polynomials:sec},
we focus on \emph{polynomials}, by introducing the \cref{karh:thm} of
\textcite{KARH77}, providing a counter-example in \cref{thm:counter-example},
and an alternative characterization in \cref{corrected-version:thm}. In
\cref{beyond-polynomials:sec}, we leverage the results over polynomials to
decide $\NPoly$ inside $\ZPoly$ in the \kl{commutative} case
(\cref{decidable-n-poly:thm}), and prove that \cref{zsf-nsf:conjecture} holds
for \kl{commutative} functions (\cref{zsf-npoly-nsf:thm}). Finally, in
\cref{beyond-commutative:sec}, we introduce the notion of \kl{residual
transducer} (\cref{residual-transducer:def}), and prove in
\cref{non-commutative-npoly:thm} that it characterizes $\NPoly$ inside
$\ZPoly$.

>>>>

There have been recurrent attempts to generalize the notions of regularity from
languages to functions, leading to classes of various expressiveness
(sequential, rational, regular, and \kl{polyregular functions}). In this paper,
we will focus on \kl{polyregular functions} ($\Poly$), that shares with regular
languages its characterizations in terms of logic (via $\MSO$ interpretations),
automata (via pebble automata), and adds new bridges with programming languages
via its relationships with lambda-terms and so-called for-transducers
\cite{bojanczyk2019string}. As for regular languages, the class
$\SF$ of \intro{star-free polyregular function} can equivalently be introduced
in terms of first-order definability, counter-free pebble transducers, and
suitable restrictions of for-tranducers.

Classical problems, including the decidability of aperiodicity, become
difficult in the case of \kl{polyregular functions}, due to the lack of
canonical objects akin to the syntactic monoid or the minimal automaton.
Indeed, for sequential functions that admit a canonical object, aperiodicity is
decidable \cite{choffrut03}, and more recently, aperiodicity has been proven
decidable for rational functions
\cite{filiot2016aperiodicity,filiot2018canonical}. The problem remains open for
regular functions, although it has been proven decidable when changing the
semantics of function equality to account for the \emph{origins} in the output
\cite{bojanczyk14}. To construct subclasses of \kl{polyregular function}, an
interesting parameter is the \intro{growth rate} of a function $f$, which is given
by a number $k \in \Nat$, which can be used to create a strict hierarchy inside
$\Poly$. We write $\Poly[k]$ for the \kl{polyregular functions} of \kl{growth
rate} $k$. It has been shown that the parameter $k$ controls the maximal
complexity of the $\MSO$ interpretation needed to represent the function
\cite{bojanczyk2022transducers}.

@@

Recently, aperiodicity has been shown decidable over a restricted class of
\kl{polyregular functions} where the ouptut belongs to $\Rel$ instead of
$\Nat$, that are called \kl{$\Rel$-polyregular functions} ($\ZPoly$)
\cite{LOPEZ23b}. These functions generalize \kl{$\Nat$-polyregular functions}
by allowing negative outputs, and therefore enabling a limited form of
backtracking. As for the unary outputs, these results generalize to
\kl{polyregular functions} having outputs in the free commutative \emph{group}
generated by a finite alphabet $\Gamma$. The decidability of aperiodicity for
\kl{$\Rel$-polyregular functions} relies on the construction of a canonical
object called the \emph{residual transducer}, the latter being essentially
based on differences between functions, crucially leveraging negative outputs
(i.e., the presence of inverse elements). If the proof method does not carry
from $\Rel$-output functions to $\Nat$-output functions, it was conjectured
that the semantic characterization of \kl{star-free $\Rel$-polyregular
functions} ($\ZSF$) would continue to characterize \kl{star-free
$\Nat$-polyregular functions} ($\NSF$).

@@

It is well known that $\NPoly$ (resp. $\ZPoly$) is proper subclass respectively
of \kl{$\Nat$-rational series} ($\NRat$) (resp. of \kl{$\Rel$-rational series}
$\ZRat$), that have been characterized in terms of \kl{growth rate}, ambiguity,
and a notion of \emph{quantitative} $\MSO$
\cite{schutzenberger1962,kreutzer2013}, and therefore can benefit from generic
theorems obtained on these \cite{berstel2011noncommutative}. However, for these
larger classes of functions, several competing notions of aperiodicity exist
\cite{droste2019aperiodic,reutenauer_series_1980,LOPEZ23b}, and the
decidability of the ones that extends $\ZSF$ and $\NSF$ remains open.
Furthermore, the decidability of $\NRat$ inside $\ZRat$ is also a longstanding
open problem, that was to the knowledge of the authors, only solved by
\textcite{KARH77} in the case of \kl{$\Rel$-rational series} that compute
\emph{polynomials}, with an incorrect proof. The decidability of $\NPoly$
inside $\ZPoly$ (resp. $\NSF$ inside $\ZSF$) remains an open problem \cite[Open
question 5.55]{gaetanphd}.

@@

\begin{figure}
    \includestandalone[width=5cm]{tikz/class-inclusions}
    \caption{
        Decidability and inclusions of classes of functions,
        arranged along two axes. The first one is the complexity
        of the output alphabet ($\Nat$, $\Rel$, $\Sigma$). The second
        one is the allowed computational power
        (star-free polyregular functions, polyregular functions, 
        rational series).
    }
    \label{previously-known-inclusions:fig}
\end{figure}

\paragraph*{Contributions.} This paper focuses on \kl{$\Nat$-polyregular
functions} that are furthermore \kl{commutative}. Remark that this enforces a
commutativity property both on the input alphabet and on the output alphabet.
In this restricted setting, we answer positively to \cref{zsf-nsf:conjecture},
and prove that the membership problems between $\NSF$, $\ZSF$, $\NPoly$, and
$\ZPoly$ are all decidable, with effective conversion algorithms. This
effectively shows that all arrows in the lower left square of
\cref{previously-known-inclusions:fig} are decidable in the \kl{commutative}
setting. Furthermore, we characterize \kl{commutative} \kl{$\Nat$-polyregular
function} as the class of \kl{$\Rel$-polyregular functions} that are
\kl{combinatorial} (defined page \kpageref{combinatorial}), which is a semantic
property akin to the one \kl{ultimate polynomiality} that characterizes $\ZSF$.
As a byproduct of our analysis, we provide a counter example to the
characterization of polynomials computed by \kl{$\Nat$-rational series} given
by \textcite[Theorem 3.3, page 4; re-introduced as \cref{karh:thm}]{KARH77}.

Finally, we make a first step towards understanding the non-commutative by
introducing a canonical model of computation for \kl{$\Nat$-polyregular
functions} built upon residuals, called the \kl{residual transducer}. This
allows the characterization of $\NPoly[k]$ as the class of functions $f$ such
that the collection $\Res(f)$ of all \kl{residuals} of $f$ is a
\kl{well-quasi-ordering}, generalizing the characterization of $\ZPoly[k]$ as
the class of functions $f$ such that $\Res(f)$ has finite index (for some
equivalence relation). The \kl{residual transducer} is effectively computable
if $f$ is a \kl{commutative} \kl{$\Nat$-polyregular function}, and it is
conjectured to be computable for all functions in $\NPoly$.

\paragraph*{Outline of the paper.} In \cref{preliminaries:sec}, we introduce
the multiple characterizations of \kl{$\Nat$-polyregular function} (resp.
$\Rel$-polyregular functions), show that one can decide if a function $f \in
\ZPoly$ is \kl{commutative} (\cref{decidable-commutative-poly:lemma}), and give the
basic definitions used on multivariate polynomials. In \cref{polynomials:sec},
we focus on \emph{polynomials}, by introducing the \cref{karh:thm} of
\textcite{KARH77}, providing a counter-example in \cref{thm:counter-example},
and an alternative characterization in \cref{corrected-version:thm}. In
\cref{beyond-polynomials:sec}, we leverage the results over polynomials to
decide $\NPoly$ inside $\ZPoly$ in the \kl{commutative} case
(\cref{decidable-n-poly:thm}), and prove that \cref{zsf-nsf:conjecture} holds
for \kl{commutative} functions (\cref{zsf-npoly-nsf:thm}). Finally, in
\cref{beyond-commutative:sec}, we introduce the notion of \kl{residual
transducer} (\cref{residual-transducer:def}), and prove in
\cref{non-commutative-npoly:thm} that it characterizes $\NPoly$ inside
$\ZPoly$.
